Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-02 latex/main.tex
Updates from Overleaf

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and
viral relationships can be represented by a single tree.
However, important SARS-CoV-2 recombinant lineages have emerged.
A more suitable approach is to use ancestral recombination graphs (ARG),
which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
Recent computational advances have made ARG inference possible for datasets of millions of genome sequences.
Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and
reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences.
Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023.
We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution —
the backbone phylogeny, mutational spectra, and major recognized recombinants.
This ARG enables us to explore the evolutionary origins of known and novel recombinants.
Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and
can be analyzed efficiently and interactively using standard Python tools on a typical laptop.
This combination of capabilities makes sc2ts a powerful tool
for studying the recombination and mutation processes driving SARS-CoV-2 evolution.

@@

Using sc2ts, we built an ARG of ~2.48 million-samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM (~5 CPU years; Figure SX).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as “robust” recombination events, Figure SX; see Document S1),
reflecting the overwhelming tree-like structure of SARS-CoV-2 evolutionary history.

@@

We present sc2ts, a real-time ARG inference method,
and applied it to build a \~2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
By identifying specific samples as parents,
sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods,
which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages.
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022}
which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
Combined with new methods and tools to evaluate evidence for recombination,
sc2ts is a significant technological advance over current methods for the discovery and
analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
We hope that the ARG will spur on new research areas,
in which recombination is naturally and fully accounted for,
and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
has made on our understanding of human genetic ancestry and demographic past (e.g., REF).

>>>>

%%% The summary (abstract) should consist of a single paragraph of 150 words or fewer.
%%% This is the SMBE abstract. We can rewrite or work off of it, but it needs to be shorter.
Widely used phylogenetic methods to study SARS-CoV-2 evolution assume that genetic recombination is negligible and
viral relationships can be represented by a single tree.
However, important SARS-CoV-2 recombinant lineages have emerged.
A more suitable approach is to use ancestral recombination graphs (ARGs),
which jointly model recombination, mutation, and genetic inheritance in a cohesive framework.
Recent computational advances have made ARG inference possible for datasets of millions of genome sequences.
Building on these new advances, we have developed sc2ts, a scalable method to simultaneously detect recombinants and
reconstruct ARGs from heterochronous SARS-CoV-2 genome sequences.
Using sc2ts, we inferred an ARG of 2.48 million genomes sampled from January, 2020 to February, 2023.
We find that this pandemic-scale ARG unifies known features of SARS-CoV-2 evolution —
the backbone phylogeny, mutational spectra, and major recognized recombinants.
This ARG enables us to explore the evolutionary origins of known and novel recombinants.
Furthermore, the ARG is compactly stored in the tree sequence format using the tskit library and
can be analyzed efficiently and interactively using standard Python tools on a typical laptop.
This combination of capabilities makes sc2ts a powerful tool
for studying the recombination and mutation processes driving SARS-CoV-2 evolution.

@@

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM (~5 CPU years; Figure SX).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have a minimum of four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as “robust” recombination events, Figure SX; see Document S1),
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

We present sc2ts, a real-time ARG inference method,
and applied it to build a $\sim$2.48 million-sample pandemic-scale ARG of SARS-CoV-2.
By identifying specific samples as parents,
sc2ts enables more precise interpretation of recombinants than state-of-the-art recombination detection methods,
which find recombinants by matching a given sample to pre-defined population-averaged mutational profiles of Pango lineages.
Sc2ts treats recombination as an integral part of SARS-CoV-2 evolution, in contrast to RIPPLES \cite{Turakhia2022}
which detects recombinants post hoc by searching for unusual branches in a given phylogeny,
inevitably resulting in an incomprehensive representation of SARS-CoV-2 evolution.
Combined with new methods and tools to evaluate evidence for recombination,
sc2ts is a significant technological advance over current methods for the discovery and
analysis of recombinants and efficient inference of SARS-CoV-2 genealogy.

The ARG itself is a useful resource to support future studies of SARS-CoV-2.
Our results show that the ARG captures and unifies important features of SARS-CoV-2 evolution,
which are shown in different studies using different methods, tools, and approaches \cite{Turakhia2022,Bloom2023,Hunt2024}.
The ARG displays remarkable phylogenetic congruence with a state-of-the-art phylogeny built using UShER.
Also, the ARG compactly encodes genetic variation in millions of SARS-CoV-2 genomes,
allowing for convenient and efficient reuse of this valuable dataset
for a variety of evolutionary and genetic analyses beyond those presented here.
% We hope that the ARG will spur on new research areas,
% in which recombination is naturally and fully accounted for,
% and will make an impact comparable to the impact that a recently published human ARG \cite{Wohns2022}
% has made on our understanding of human genetic ancestry and demographic past (e.g., REF).
So far, recent developments in ARG methods have focused almost 
exclusively on applications to human sequencing data. In that space,
the availability of ARGs reconstructed from large datasets has already
had tremendous impact on improving our understanding of human genetic
ancestry and past demography \cite{Wohns2022}. We expect
similar leaps to be enabled through tailoring ARG reconstruction methods
to data from other species and pathogens. Specifically, existing
inference methods can be adapted to use the sc2ts ARG for the 
genealogy-based study of selection \cite{hejase2022deep} or the
geographic spread of the virus through time
\cite{grundler2025geographic}. This is in addition to simpler
analyses, such as calculating mutation spectra or querying the number
or frequency of mutations at a given locus, which the ARG makes 
significantly more efficient, and ultimately more easily reproducible.
Since the ARG is a condensed representation of the genetic information
contained in the sequencing data, it can also be used to understand,
for instance, how the variability in sequencing efforts across time 
and location affects our ability to reconstruct the origin of 
important lineages and detect new recombinants.
Since reconstructing the ARG is computationally expensive, 
and checking and improving various aspects of its quality is laborious, 
we hope that making this resource openly available will save
significant time and 
effort for other researchers and enable immediate focus on analysing 
the information contained within this data. In-depth analysis of the
ARG is also incredibly useful for identifying important directions for
further research, both regarding methods (where current models do not
correctly capture features of the data) and data processing 
(where gaps in sequencing or data artifacts result in clear
reconstruction errors).
